Imputation of Missing Values in the Melbourne CBD Pedestrian Dataset

### 2017-09-03 thesis/missing_impute.Rmd
model for flinders st - swanston st (west) using neighbouring sensors

---
title: "Imputation of Missing Values in the Melbourne CBD Pedestrian Dataset"
output:
  html_document:
    theme: united
  html_notebook: default
  pdf_document: default
---
```{r, include=FALSE}
library(plyr)
library(tidyverse)
library(lubridate)
library(magrittr)
library(viridis)
library(gridExtra)
set.seed(12345)
```

@@

ped_coeffs <- list()
ped_model_fitted <- list()
ped_model_summary <- list()
p <- progress_estimated(43)
options(na.action = 'na.omit')
for (i in 1:43) {
  p$tick()$print()
  train <- ped_list[[i]]
  ped_model_i <- glm(Hourly_Counts ~ Month + DayType*Time,
                     family = quasipoisson(), data = train)
  ped_coeffs[[i]] <- ped_model_i$coefficients
  ped_model_summary[[i]] <- summary(ped_model_i)
}
rm(ped_model_i)

```
The code above fits a GLM model for each sensor location. This model uses a quasi-poisson distribution to allow of overdispersion. Using a standard poisson distribution would give the restriction of mean = variance.  
The model is specified as `Hourly_Counts ~ Month + DayType * Time`, where `Month` is given an additive effect, and `DayType * Time` gives an interaction term between `DayType` and `Time`, as well as any other indirect additive effects.
To improve efficiency, only the coefficients of the model are retained, as well as model summarries for model evaluation.  
We train the model only on the data from 2014 to 2016. The test data set is the data available from 2017.

@@

for (i in 1:43) {
  test_df <- model.matrix(Hourly_Counts ~ Month + DayType*Time,
                          data = test_ped[[i]])
  test_coeff <- ped_coeffs[[i]]
  if (length(test_coeff) < ncol(test_df)) {
    missing_cols <- setdiff(colnames(test_df), rownames(as.data.frame(test_coeff)))
    test_df <- test_df[, !(colnames(test_df) %in% missing_cols)]
  }

  test_preds <- exp(test_df %*% test_coeff)
  test_ped[[i]]$Fitted <- test_preds
}

@@

# Goodness of Fit
### Visually
```{r, fig.align= 'center', fig.height=24, fig.width=10, fig.retina=1, cache=TRUE, echo=FALSE}
ggplot(filter(test_fitted, Date > "2017-07-23")) +
  geom_line(aes(x = Date_Time, y = Hourly_Counts)) +
  geom_line(aes(x = Date_Time, y = Fitted),
            colour = "violetred2", alpha = 0.75) +
                         facet_wrap(~ Sensor, scales = "free_y", ncol = 4) + theme_minimal() + 
                    theme(axis.text.x = element_text(size = 3),
                          strip.text = element_text(size = 10),
                          axis.text.y = element_text(size = 0)) +
                    xlab("Date") + ylab("Hourly Counts")
```
Looking at plots of actuals (black) against fitted (red) of data from 24 July 2017 to 6 August 2017, we see that some sensors appear to be not operational. We see that there is a mix of very good fitting models and not so good ones. An example of of a reasonably good fitting model is the _Collins Place (North)_ sensor location:
```{r, fig.align= 'center', fig.retina = 1, cache=TRUE}
ggplot(test_fitted %>%
         filter(Date > "2017-07-23",
                Sensor == "Collins Place (North)")) +
      geom_line(aes(x = Date_Time, y = Hourly_Counts)) +
      geom_line(aes(x = Date_Time, y = Fitted), colour = 'violetred2', alpha = 0.7) +
      theme_minimal()
```

@@

```{r, fig.align= 'center', fig.retina=1, echo=FALSE, cache=TRUE, warning=F}
ggplot(filter(test_fitted, Sensor_ID == "Grattan St-Swanston St (West)")) +
  geom_path(aes(x = Time, y = Hourly_Counts, group = Date), alpha = 0.1) +
  geom_path(aes(x = Time, y = Fitted, group = Date), colour = "violetred2", alpha = 0.1) +
  facet_wrap(~ DayType) +
      theme_minimal()
```

@@

>>>>

---
title: "Imputation of Missing Values in the Melbourne CBD Pedestrian Dataset"
output:
  html_document:
    theme: united
  html_notebook: default
  pdf_document: default
---
```{r, include=FALSE}
library(plyr)
library(tidyverse)
library(lubridate)
library(pander)
library(gridExtra)
library(fortify)
set.seed(12345)
```

@@

# ped_coeffs <- list()
# ped_model_fitted <- list()
# ped_model_summary <- list()
# p <- progress_estimated(43)
# options(na.action = 'na.omit')
# for (i in 1:43) {
#   p$tick()$print()
#   train <- ped_list[[i]]
#   ped_model_i <- glm(Hourly_Counts ~ Month + DayType*Time,
#                      family = quasipoisson(), data = train)
#   ped_coeffs[[i]] <- ped_model_i$coefficients
#   ped_model_summary[[i]] <- summary(ped_model_i)
# }
# rm(ped_model_i)
```

```{r, cache = T}
## Alternative code for GLM
library(foreach)
library(doSNOW)
### code from:
## http://www.win-vector.com/blog/2014/05/trimming-the-fat-from-glm-models-in-r/
stripGlmLR = function(cm) {
  cm$y = c()
  cm$model = c()

  cm$residuals = c()
  cm$fitted.values = c()
  cm$effects = c()
  cm$qr$qr = c()  
  cm$linear.predictors = c()
  cm$weights = c()
  cm$prior.weights = c()
  cm$data = c()

  cm$family$variance = c()
  cm$family$dev.resids = c()
  cm$family$aic = c()
  cm$family$validmu = c()
  cm$family$simulate = c()
  attr(cm$terms,".Environment") = c()
  attr(cm$formula,".Environment") = c()

  cm
}
#######################################

cl <- makeCluster(2)
registerDoSNOW(cl)
ped_models <- foreach(i = 1:43) %dopar% {
  model <- glm(Hourly_Counts ~ Month + DayType*Time, data = ped_list[[i]], family = quasipoisson())
  model <- stripGlmLR(model)
  return(model)
}

ped_models[[31]]$xlevels <- ped_models[[1]]$xlevels

```

The code above fits a GLM model for each sensor location. This model uses a quasi-poisson distribution to allow of overdispersion. Using a standard poisson distribution would give the restriction of mean = variance.  
The model is specified as `Hourly_Counts ~ Month + DayType * Time`, where `Month` is given an additive effect, and `DayType * Time` gives an interaction term between `DayType` and `Time`, as well as any other indirect additive effects.
To improve efficiency, only the coefficients of the model are retained, as well as model summarries for model evaluation.  
We train the model only on the data from 2014 to 2016. The test data set is the data available from 2017.

@@

for (i in 1:43) {
  test_ped[[i]]$Fitted <- predict.glm(ped_models[[i]], newdata = test_ped[[i]], type = "response")
}

# 
# for (i in 1:43) {
#   test_df <- model.matrix(Hourly_Counts ~ Month + DayType*Time,
#                           data = test_ped[[i]])
#   test_coeff <- ped_coeffs[[i]]
#   if (length(test_coeff) < ncol(test_df)) {
#     missing_cols <- setdiff(colnames(test_df), rownames(as.data.frame(test_coeff)))
#     test_df <- test_df[, !(colnames(test_df) %in% missing_cols)]
#   }
#   
#   test_preds <- exp(test_df %*% test_coeff)
#   test_ped[[i]]$Fitted <- test_preds
# }

@@

# Goodness of Fit
### Visually
```{r, fig.align= 'center', fig.height=24, fig.width=10, fig.retina=2, cache=TRUE, echo=FALSE}
ggplot(filter(test_fitted, Date > "2017-07-23")) +
  geom_line(aes(x = Date_Time, y = Hourly_Counts)) +
  geom_line(aes(x = Date_Time, y = Fitted),
            colour = "violetred2", alpha = 0.75) +
         facet_wrap(~ Sensor, scales = "free_y", ncol = 4) + theme_minimal() + 
    theme(axis.text.x = element_text(size = 3),
          strip.text = element_text(size = 10),
          axis.text.y = element_text(size = 0)) +
    xlab("Date") + ylab("Hourly Counts") +
    ggtitle("Plot of Actual (black) against Fitted for period 24 July 2017 to 6 August 2017")        
```
Looking at plots of actuals (black) against fitted of data from 24 July 2017 to 6 August 2017, we see that some sensors appear to be not operational. We see that there is a mix of very good fitting models and not so good ones. An example of of a reasonably good fitting model is the _Collins Place (North)_ sensor location:
```{r, fig.align= 'center', fig.retina = 1, cache=TRUE}
ggplot(test_fitted %>%
         filter(Date > "2017-07-23",
                Sensor == "Collins Place (North)")) +
      geom_line(aes(x = Date_Time, y = Hourly_Counts)) +
      geom_line(aes(x = Date_Time, y = Fitted), colour = 'violetred2', alpha = 0.7) +
      theme_minimal()
```

@@

```{r, fig.align= 'center', fig.retina=1, echo=FALSE, cache=TRUE, warning=F}
ggplot(filter(test_fitted, Sensor_ID == "Grattan St-Swanston St (West)")) +
  geom_path(aes(x = as.integer(Time), y = Hourly_Counts, group = Date), alpha = 0.1) +
  geom_path(aes(x = as.integer(Time), y = Fitted, group = Date), colour = "violetred2", alpha = 0.1) +
  facet_wrap(~ DayType) +
      theme_minimal()
```

@@

Computing the RMSE for all of the test set (2017 data) for each sensor:

```{r}
test_df <- test_fitted %>% mutate(residsq = (Fitted - Hourly_Counts)^2) %>%
            split(.$Sensor_ID)
test_df_rmse <- data.frame(Sensor_ID = rep(0,43), RMSE = rep(0, 43), StdRMSE = rep(0, 43))
  for (i in 1:43) {
      test_df_rmse$Sensor_ID[i] <- names(test_df)[i]
      test_df_rmse$RMSE[i] <- mean(test_df[[i]]$residsq) %>% sqrt()
      test_df_rmse$StdRMSE[i] <- test_df_rmse$RMSE[i] / mean(test_df[[i]]$Hourly_Counts)

  }

test_df_plot <- test_fitted %>% filter(Date > "2017-07-23") %>%
            mutate(residsq = (Fitted - Hourly_Counts)^2) %>%
            split(.$Sensor_ID)
test_df_rmse_plot <- data.frame(Sensor_ID = rep(0,43), RMSE = rep(0, 43), StdRMSE = rep(0, 43))
  for (i in 1:43) {
      test_df_rmse_plot$Sensor_ID[i] <- names(test_df_plot)[i]
      test_df_rmse_plot$RMSE[i] <- mean(test_df_plot[[i]]$residsq) %>% sqrt()
      test_df_rmse_plot$StdRMSE[i] <- test_df_rmse_plot$RMSE[i] / mean(test_df_plot[[i]]$Hourly_Counts)
  }
pander(arrange(test_df_rmse, by= StdRMSE))
```

Here, the `StdRMSE` is the RMSE standardised by the mean.

```{r}
library(ggmap)
  ped_loc <- read_csv("../data/Pedestrian_sensor_locations.csv")
```

Looking at some problematic sensor locations:
```{r}
pander(test_df_rmse[test_df_rmse$StdRMSE > 0.65, ] %>% arrange(by = StdRMSE))
test_df_rmse[test_df_rmse$StdRMSE > 0.65, ] %>% .$Sensor_ID -> bad_sensors
```

```{r, warning=F}
p1 <- ggplot(test_ped$`Birrarung Marr`) + geom_path(aes(x = Time, y = Hourly_Counts, group = Date), alpha = 0.05) +
                                  geom_path(aes(x = Time, y = Fitted, group = Date), alpha = 0.05,
                                            colour = "violetred2") + ggtitle("Birrarung Marr")
p2 <- ggplot(test_ped$`Birrarung Marr`) + geom_line(aes(x = Date_Time, y = Hourly_Counts), alpha = 0.5) +
                                  geom_line(aes(x = Date_Time, y = Fitted), alpha = 0.5, colour = "violetred2")

grid.arrange(p1, p2, nrow = 1)
```

```{r}
p1 <- ggplot(test_ped$`Flagstaff Station`) + geom_path(aes(x = Time, y = Hourly_Counts, group = Date), alpha = 0.05) +
                                  geom_path(aes(x = Time, y = Fitted, group = Date), alpha = 0.05,
                                            colour = "violetred2") + ggtitle("Flagstaff Station")
p2 <- ggplot(test_ped$`Flagstaff Station`) + geom_line(aes(x = Date_Time, y = Hourly_Counts), alpha = 0.5) +
                                  geom_line(aes(x = Date_Time, y = Fitted), alpha = 0.5, colour = "violetred2")
grid.arrange(p1, p2, nrow = 1)

```

```{r}
p1 <- ggplot(test_ped$`Flinders St-Swanston St (West)`) + geom_path(aes(x = Time, y = Hourly_Counts, group = Date), alpha = 0.05) +
                                  geom_path(aes(x = Time, y = Fitted, group = Date), alpha = 0.05,
                                            colour = "violetred2") + ggtitle("Flinders St-Swanston St (West)")
p2 <- ggplot(test_ped$`Flinders St-Swanston St (West)`) + geom_line(aes(x = Date_Time, y = Hourly_Counts), alpha = 0.5) +
                                  geom_line(aes(x = Date_Time, y = Fitted), alpha = 0.5, colour = "violetred2")
grid.arrange(p1, p2, nrow = 1)

```

```{r}
p1 <- ggplot(test_ped$`New Quay`) + geom_path(aes(x = Time, y = Hourly_Counts, group = Date), alpha = 0.05) +
                                  geom_path(aes(x = Time, y = Fitted, group = Date), alpha = 0.05,
                                  colour = "violetred2") + ggtitle("New Quay")
p2 <- ggplot(test_ped$`New Quay`) + geom_line(aes(x = Date_Time, y = Hourly_Counts), alpha = 0.5) +
                                  geom_line(aes(x = Date_Time, y = Fitted), alpha = 0.5, colour = "violetred2")
grid.arrange(p1, p2, nrow = 1)

```

```{r}
p1 <- ggplot(test_ped$`Princes Bridge`) + geom_path(aes(x = Time, y = Hourly_Counts, group = Date), alpha = 0.05) +
                                  geom_path(aes(x = Time, y = Fitted, group = Date), alpha = 0.05,
                                  colour = "violetred2") + ggtitle("Princes Bridge")
p2 <- ggplot(test_ped$`Princes Bridge`) + geom_line(aes(x = Date_Time, y = Hourly_Counts), alpha = 0.5) +
                                  geom_line(aes(x = Date_Time, y = Fitted), alpha = 0.5, colour = "violetred2")
grid.arrange(p1, p2, nrow = 1)
```

```{r}
p1 <- ggplot(test_ped$`Southbank`) + geom_path(aes(x = Time, y = Hourly_Counts, group = Date), alpha = 0.05) +
                                  geom_path(aes(x = Time, y = Fitted, group = Date), alpha = 0.05,
                                  colour = "violetred2") + ggtitle("Southbank")
p2 <- ggplot(test_ped$`Southbank`) + geom_line(aes(x = Date_Time, y = Hourly_Counts), alpha = 0.5) +
                                  geom_line(aes(x = Date_Time, y = Fitted), alpha = 0.5, colour = "violetred2")
grid.arrange(p1, p2, nrow = 1)
```

### Standardise Counts

$$HourlyCounts_{i,j} \sim N(0,1)$$

```{r}
for (i in 1:43) {
  ped_list[[i]]$scaled_count <- scale(ped_list[[i]]$Hourly_Counts)
}
ped_df <- bind_rows(ped_list)
ped_df %>% mutate(Timep = as.integer(Time)) -> ped_df
```

```{r}
ggplot(ped_df[ped_df$Date > "2015-01-01", ]) + 
  geom_path(aes(x = Timep, y = scaled_count, colour = Sensor_ID, group = Date), alpha = 0.01) +
      theme(legend.position = 'none') + xlab("Time") + ylab("Standardised Counts")
```

```{r}
ggplot(ped_df[ped_df$Sensor_ID == "Flinders St-Swanston St (West)" | 
              ped_df$Sensor_ID == "Flinders St-Elizabeth St (East)" |
              ped_df$Sensor_ID == "Flinders St Station Underpass" , ]) +
      geom_smooth(aes(x = Timep, y = scaled_count, colour = Sensor_ID))

ggplot(ped_df[ped_df$Sensor_ID == "Flinders St-Swanston St (West)" | 
              ped_df$Sensor_ID == "Flinders St-Elizabeth St (East)" |
              ped_df$Sensor_ID == "Flinders St Station Underpass" , ]) +
      geom_path(aes(x = Timep, y = scaled_count, group = Date, colour = Sensor_ID),
                alpha = 0.05) + facet_wrap(~ DayType) + theme(legend.position = 'none')
```

```{r}
ped_df[ped_df$Sensor_ID == "Flinders St-Swanston St (West)", ]$scaled_count -> swan
ped_df[ped_df$Sensor_ID == "Flinders St-Elizabeth St (East)", ]$scaled_count -> liz
ped_df[ped_df$Sensor_ID == "Flinders St Station Underpass", ]$scaled_count -> underpass

flin_swan_df <- data.frame(swan = swan, liz = liz, underpass = underpass,
                           date = ped_df[ped_df$Sensor_ID == "Webb Bridge", ]$Date,
                           time = as.integer(ped_df[ped_df$Sensor_ID == "Webb Bridge", ]$Time),
                           timefact = ped_df[ped_df$Sensor_ID == "Webb Bridge", ]$Time,
                           month = ped_df[ped_df$Sensor_ID == "Webb Bridge", ]$Month,
                           daytype = ped_df[ped_df$Sensor_ID == "Webb Bridge", ]$DayType,
                           dtm = ped_df[ped_df$Sensor_ID == "Webb Bridge", ]$Date_Time,
                           hday = ped_df[ped_df$Sensor_ID == "Webb Bridge", ]$IsHDay)

flin_swan <- lm(swan ~ liz + underpass, data = flin_swan_df) 
summary(flin_swan)
flin_swan_df$swan_fit <- predict.lm(flin_swan, newdata = flin_swan_df)
flin_swan_df$resid <- flin_swan_df$swan_fit - flin_swan_df$swan
plot(flin_swan)
ggplot(flin_swan_df) + 
  geom_path(aes(x = time, y = swan, group = date), alpha = 0.05) + 
  geom_path(aes(x = time, y = liz, group = date), alpha = 0.02, colour = 'limegreen') + 
  geom_path(aes(x = time, y = swan_fit, group = date), alpha = 0.02, colour = 'red')

```

```{r}
flin_swan_df %>% mutate(midday = ifelse(time > 10 & time < 17, 1, 0)) -> flin_swan_df

flin_swan <- lm(swan ~ daytype*liz*timefact, data = flin_swan_df) 
summary(flin_swan)
flin_swan_df$swan_fit <- predict.lm(flin_swan, newdata = flin_swan_df)
flin_swan_df$resid <- flin_swan_df$swan_fit - flin_swan_df$swan
plot(flin_swan)
ggplot(flin_swan_df) + 
  geom_path(aes(x = time, y = swan, group = date), alpha = 0.05) + 
  geom_path(aes(x = time, y = liz, group = date), alpha = 0.02, colour = 'limegreen') + 
  geom_path(aes(x = time, y = swan_fit, group = date), alpha = 0.02, colour = 'red')

```

```{r}
flin_swan_2 <- lm(swan ~ daytype*timefact*liz + daytype*timefact*underpass, data = flin_swan_df) 
summary(flin_swan_2)
flin_swan_df$swan_fit <- predict.lm(flin_swan_2, newdata = flin_swan_df)
flin_swan_df$resid <- flin_swan_df$swan_fit - flin_swan_df$swan
plot(flin_swan_2)

ggplot(flin_swan_df) + 
  geom_path(aes(x = time, y = swan, group = date), alpha = 0.05) + 
  geom_path(aes(x = time, y = liz, group = date), alpha = 0.02, colour = 'limegreen') + 
  geom_path(aes(x = time, y = swan_fit, group = date), alpha = 0.02, colour = 'red')

```

Issues with potentially `NA` having `Hourly_Counts = 0`:

```{r}
ggplot(flin_swan_df) + geom_line(aes(x = dtm, y = swan))
ggplot(ped_data %>% filter(Sensor_ID == "Flinders St-Swanston St (West)")) + geom_line(aes(x = Date_Time, y = Hourly_Counts))
```

Need to filter out pre-2015 data from training set:

```{r}
flin_swan_df_train <- flin_swan_df %>% filter(date > "2014-12-31")

flin_swan_2 <- lm(swan ~ timefact*liz + timefact*underpass, data = flin_swan_df_train) 
summary(flin_swan_2)
flin_swan_df$swan_fit <- predict.lm(flin_swan_2, newdata = flin_swan_df)
flin_swan_df$resid <- flin_swan_df$swan_fit - flin_swan_df$swan
plot(flin_swan_2)

ggplot(flin_swan_df) + 
  geom_path(aes(x = time, y = swan, group = date), alpha = 0.05) + 
  geom_path(aes(x = time, y = liz, group = date), alpha = 0.02, colour = 'limegreen') + 
  geom_path(aes(x = time, y = swan_fit, group = date), alpha = 0.02, colour = 'red')

```
Also removed `DayType` due to high leverage observations.

```{r}
flin_swan_df[8763:8764, ]
```

High leverage points here are the New Year's Day celebrations.
